---
title: 'Storms: Health and Economic Impacts'
output:
  html_document:
    keep_md: yes
  pdf_document: default
date: "September 2014"
---

<br>

## Synopsis

This report explores the social and economic impact of storm events since 1996 to
2011 across USA.

Starting with data from NOAA storm database, there is a deep data
clean process to adjust data and events to the description in the codebook. It
has been a hard work to complete this part because raw data contains lot of
coding errors that are outside the codebook. Moreover, data used has been
filtered between 1996 and 2011: previous data is incomplete and even more error prone.

The report concludes with the top 10 events that are most harmful with respect
to population, both fatalities and injuries, and with respect to economic impact.
Surprisingly, the event that causes more fatalities is *excessive heat*.

<br>

## Data Processing

This section contains all preliminar data processing. It consist mainly in the
filtering, cleaning and management of raw data to get a *tidy data set*.

Libraries used during this process, and later, are:

```{r config, cache=TRUE}
library(ggplot2)
library(plyr)
library(stringr)
```

<br>

### Downloading Data

First at all, data can be downloaded from the following URL:

```{r download-data, cache=TRUE}
if(! file.exists("StormData.csv.bz2")) {
    url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(url, destfile="StormData.csv.bz2", method="curl")
}
```

<br>

### Reading data

We first begin by reading the complete *raw* data set:

```{r read-data, cache=TRUE}
StormData.raw <- read.csv("StormData.csv.bz2", header=TRUE)
```

<br>

### Filtering by year

According to the
[NOAA description of Storm Events Database Contents](http://www.ncdc.noaa.gov/stormevents/details.jsp), 
it is since 1996 that all storms events are stored in the database. We will use
storm data recorded since 1996 to address our report.  Before 1996 only *few part*
of stroms were reported. We are interested in all kind of storms, so taking a long
period of time without complete data is not a good point.

The raw data set filtered by year:

```{r filter-by-year, cache=TRUE}
StormData.raw.1996_1999 <- grep("19[9][6-9]", StormData.raw$BGN_DATE)
StormData.raw.2000_ <- grep("20[0-9][0-9]", StormData.raw$BGN_DATE)
StormData.raw <- StormData.raw[c(StormData.raw.1996_1999, StormData.raw.2000_),]
```

### Removing rows without interest

To address the questions about the most harmful events with respect population
and economic consequences, we will use only those events where `FATALITIES`,
`INJURIES`, `PROPDMG` or `CROPDMG` contain information (are positive).

The rest of the events are useless for the purpose of this report:

```{r removing-rows, cache=TRUE}
StormData.raw <- subset(StormData.raw, FATALITIES>0 | INJURIES>0 | PROPDMG>0 | CROPDMG>0)
```

### Cleaning **EVTYPE** column

This column needs severe cleaning. In the current data set we have at this
stage, column `EVTYPE` contains `r length(unique(StormData.raw$EVENT))` different values.

We start by doing a *first normalization of strings*, which consists in removing
of starting and trailing spaces and converting them to lowercase:

```{r clean-EVTYPE, cache=TRUE}
StormData.raw$EVTYPE <- str_trim(tolower(StormData.raw$EVTYPE))
```

Factors in this column are larger than the 48 values specified in the codebook.

```{r EVTYPE.raw, cache=TRUE}
EVTYPE.raw <- sort(unique(StormData.raw$EVTYPE))
length(EVTYPE.raw)
```

For example, in the first 20 values shown here there are values not considered
in the codebook:

```{r EVTYPE.raw-head, cache=TRUE}
head(EVTYPE.raw, n=20)
```

Even more, there is a curious value `other` to specify accidents. For example:

```{r other, cache=TRUE}
as.character(StormData.raw[StormData.raw$EVTYPE=="other", "REMARKS"][1])
```

From this event *REMARKS* the final *EVENT* could be *drought*, *heavy snow*,
*avalanche*, *flood*, *excessive heat* or *flash flood*. Futhermore, the event
could be split into several events occurring in different places: heavy snow in
the city of Valdez, one avalanche in the Richardson Highway and another one in
a local home, and finally a flood in the city streets (Valdez?).

There are `r length(StormData.raw$EVTYPE[StormData.raw$EVTYPE=="other"])` rows
with this *EVTYPE* value, with vague descriptions or with multipe events.
We will ignore these events. This leads a number of *EVTYPE* values of:

```{r remove-EVTYPE-other, cache=TRUE}
StormData.raw <- subset(StormData.raw, EVTYPE!="other")
EVTYPE.raw <- sort(unique(StormData.raw$EVTYPE))
length(EVTYPE.raw)
```

To clean `EVTYPE` column we create a new column `EVENT` that maps current
*EVTYPE* values into valid *EVENT* values. The map has been elaborated manually,
exploring each one of the 179 *EVTYPE* values and assigning the most likely *EVENT*
value.

This is the table used to map current *EVTYPE* values to **EVENT** values:

<pre style="font-size:11px;line-height:1.0em">
| EVTYPE                    | EVENT                    |    | EVTYPE                    | EVENT                    |
|---------------------------+--------------------------|    |---------------------------+--------------------------|
| agricultural freeze       | frost/freeze             |    | lake effect snow          | lake-effect snow         |
| astronomical high tide    | storm surge/tide         |    | lake-effect snow          | lake-effect snow         |
| avalanche                 | avalanche                |    | lakeshore flood           | lakeshore flood          |
| beach erosion             | coastal flood            |    | landslide                 | debris flow              |
| black ice                 | ice storm                |    | landslides                | debris flow              |
| blizzard                  | blizzard                 |    | landslump                 | debris flow              |
| blowing dust              | dust devil               |    | landspout                 | debris flow              |
| blowing snow              | ice storm                |    | late season snow          | winter weather           |
| brush fire                | wildfire                 |    | light freezing rain       | freezing fog             |
| coastal  flooding/erosion | coastal flood            |    | light snow                | heavy snow               |
| coastal erosion           | coastal flood            |    | light snowfall            | heavy snow               |
| coastal flood             | coastal flood            |    | lightning                 | lightning                |
| coastal flooding          | coastal flood            |    | marine accident           | marine high wind         |
| coastal flooding/erosion  | coastal flood            |    | marine high wind          | marine high wind         |
| coastal storm             | marine thunderstorm wind |    | marine strong wind        | marine strong wind       |
| coastalstorm              | marine thunderstorm wind |    | marine thunderstorm wind  | marine thunderstorm wind |
| cold                      | cold/wind chill          |    | marine tstm wind          | marine thunderstorm wind |
| cold and snow             | cold/wind chill          |    | microburst                | blizzard                 |
| cold temperature          | cold/wind chill          |    | mixed precip              | sleet                    |
| cold weather              | cold/wind chill          |    | mixed precipitation       | heavy rain               |
| cold/wind chill           | cold/wind chill          |    | mud slide                 | debris flow              |
| dam break                 | flash flood              |    | mudslide                  | debris flow              |
| damaging freeze           | frost/freeze             |    | mudslides                 | debris flow              |
| dense fog                 | dense fog                |    | non tstm wind             | high wind                |
| downburst                 | thunderstorm wind        |    | non-severe wind damage    | high wind                |
| drought                   | drought                  |    | non-tstm wind             | high wind                |
| drowning                  | heavy rain               |    | rain                      | heavy rain               |
| dry microburst            | drought                  |    | rain/snow                 | sleet                    |
| dust devil                | dust devil               |    | record heat               | excessive heat           |
| dust storm                | dust storm               |    | rip current               | flood                    |
| early frost               | frost/freeze             |    | rip currents              | flood                    |
| erosion/cstl flood        | coastal flood            |    | river flood               | flood                    |
| excessive heat            | excessive heat           |    | river flooding            | flood                    |
| excessive snow            | heavy snow               |    | rock slide                | debris flow              |
| extended cold             | extreme cold/wind chill  |    | rogue wave                | high surf                |
| extreme cold              | extreme cold/wind chill  |    | rough seas                | high surf                |
| extreme cold/wind chill   | extreme cold/wind chill  |    | rough surf                | high surf                |
| extreme windchill         | extreme cold/wind chill  |    | seiche                    | seiche                   |
| falling snow/ice          | heavy snow               |    | small hail                | hail                     |
| flash flood               | flash flood              |    | snow                      | heavy snow               |
| flash flood/flood         | flash flood              |    | snow and ice              | ice storm                |
| flood                     | flood                    |    | snow squall               | heavy snow               |
| flood/flash/flood         | flash flood              |    | snow squalls              | heavy snow               |
| fog                       | dense fog                |    | storm surge               | storm surge/tide         |
| freeze                    | frost/freeze             |    | storm surge/tide          | storm surge/tide         |
| freezing drizzle          | frost/freeze             |    | strong wind               | strong wind              |
| freezing fog              | freezing fog             |    | strong winds              | strong wind              |
| freezing rain             | winter strom             |    | thunderstorm              | thunderstorm wind        |
| freezing spray            | freezing fog             |    | thunderstorm wind         | thunderstorm wind        |
| frost                     | frost/freeze             |    | thunderstorm wind (g40)   | thunderstorm wind        |
| frost/freeze              | frost/freeze             |    | tidal flooding            | coastal flood            |
| funnel cloud              | funnel cloud             |    | tornado                   | tornado                  |
| glaze                     | frost/freeze             |    | torrential rainfall       | heavy rain               |
| gradient wind             | strong wind              |    | tropical depression       | tropical depression      |
| gusty wind                | high wind                |    | tropical storm            | tropical storm           |
| gusty wind/hail           | thunderstorm wind        |    | tstm wind                 | thunderstorm wind        |
| gusty wind/hvy rain       | high wind                |    | tstm wind  (g45)          | thunderstorm wind        |
| gusty wind/rain           | thunderstorm wind        |    | tstm wind (41)            | thunderstorm wind        |
| gusty winds               | high wind                |    | tstm wind (g35)           | thunderstorm wind        |
| hail                      | hail                     |    | tstm wind (g40)           | thunderstorm wind        |
| hard freeze               | ice storm                |    | tstm wind (g45)           | thunderstorm wind        |
| hazardous surf            | high surf                |    | tstm wind 40              | thunderstorm wind        |
| heat                      | heat                     |    | tstm wind 45              | thunderstorm wind        |
| heat wave                 | heat                     |    | tstm wind and lightning   | thunderstorm wind        |
| heavy rain                | heavy rain               |    | tstm wind g45             | thunderstorm wind        |
| heavy rain/high surf      | heavy rain               |    | tstm wind/hail            | thunderstorm wind        |
| heavy seas                | high surf                |    | tsunami                   | tsunami                  |
| heavy snow                | heavy snow               |    | typhoon                   | typhoon                  |
| heavy snow shower         | heavy snow               |    | unseasonable cold         | extrem cold/wind chill   |
| heavy surf                | high surf                |    | unseasonably cold         | extrem cold/wind chill   |
| heavy surf and wind       | high surf                |    | unseasonably warm         | excessive heat           |
| heavy surf/high surf      | high surf                |    | unseasonal rain           | heavy rain               |
| high seas                 | high surf                |    | urban/sml stream fld      | flood                    |
| high surf                 | high surf                |    | volcanic ash              | volcanic ash             |
| high surf advisory        | high surf                |    | warm weather              | excessive heat           |
| high swells               | high surf                |    | waterspout                | waterspout               |
| high water                | high surf                |    | wet microburst            | waterspout               |
| high wind                 | high wind                |    | whirlwind                 | funnel cloud             |
| high wind (g40)           | high wind                |    | wild/forest fire          | wildfire                 |
| high winds                | high wind                |    | wildfire                  | wildfire                 |
| hurricane                 | hurricane (typhoon)      |    | wind                      | high wind                |
| hurricane edouard         | hurricane (typhoon)      |    | wind and wave             | high wind                |
| hurricane/typhoon         | hurricane (typhoon)      |    | wind damage               | strong wind              |
| hyperthermia/exposure     | excessive heat           |    | winds                     | high wind                |
| hypothermia/exposure      | heavy snow               |    | winter storm              | winter storm             |
| ice jam flood (minor      | flood                    |    | winter weather            | winter weather           |
| ice on road               | ice storm                |    | winter weather mix        | winter weather           |
| ice roads                 | ice storm                |    | winter weather/mix        | winter weather           |
| ice storm                 | ice storm                |    | wintry mix                | winter weather           |
| icy roads                 | ice storm                |    '---------------------------+--------------------------'
'---------------------------+--------------------------'
</pre>

From this table, the map is:

```{r EVTYPEmap, cache=TRUE}
EVTYPEmap <- data.frame(

    EVTYPE=c("agricultural freeze", "astronomical high tide", "avalanche", "beach erosion", "black ice", "blizzard",
             "blowing dust", "blowing snow", "brush fire", "coastal  flooding/erosion", "coastal erosion",
             "coastal flood", "coastal flooding", "coastal flooding/erosion", "coastal storm", "coastalstorm",
             "cold", "cold and snow", "cold temperature", "cold weather", "cold/wind chill", "dam break",
             "damaging freeze", "dense fog", "downburst", "drought", "drowning", "dry microburst", "dust devil",
             "dust storm", "early frost", "erosion/cstl flood", "excessive heat", "excessive snow", "extended cold",
             "extreme cold", "extreme cold/wind chill", "extreme windchill", "falling snow/ice", "flash flood",
             "flash flood/flood", "flood", "flood/flash/flood", "fog", "freeze", "freezing drizzle", "freezing fog",
             "freezing rain", "freezing spray", "frost", "frost/freeze", "funnel cloud", "glaze", "gradient wind",
             "gusty wind", "gusty wind/hail", "gusty wind/hvy rain", "gusty wind/rain", "gusty winds", "hail",
             "hard freeze", "hazardous surf", "heat", "heat wave", "heavy rain", "heavy rain/high surf",
             "heavy seas", "heavy snow", "heavy snow shower", "heavy surf", "heavy surf and wind",
             "heavy surf/high surf", "high seas", "high surf", "high surf advisory", "high swells", "high water",
             "high wind", "high wind (g40)", "high winds", "hurricane", "hurricane edouard", "hurricane/typhoon",
             "hyperthermia/exposure", "hypothermia/exposure", "ice jam flood (minor", "ice on road", "ice roads",
             "ice storm", "icy roads", "lake effect snow", "lake-effect snow", "lakeshore flood", "landslide",
             "landslides", "landslump", "landspout", "late season snow", "light freezing rain", "light snow",
             "light snowfall", "lightning", "marine accident", "marine high wind", "marine strong wind",
             "marine thunderstorm wind", "marine tstm wind", "microburst", "mixed precip", "mixed precipitation",
             "mud slide", "mudslide", "mudslides", "non tstm wind", "non-severe wind damage", "non-tstm wind",
             "rain", "rain/snow", "record heat", "rip current", "rip currents", "river flood", "river flooding",
             "rock slide", "rogue wave", "rough seas", "rough surf", "seiche", "small hail", "snow", "snow and ice",
             "snow squall", "snow squalls", "storm surge", "storm surge/tide", "strong wind", "strong winds",
             "thunderstorm", "thunderstorm wind", "thunderstorm wind (g40)", "tidal flooding", "tornado",
             "torrential rainfall", "tropical depression", "tropical storm", "tstm wind", "tstm wind  (g45)",
             "tstm wind (41)", "tstm wind (g35)", "tstm wind (g40)", "tstm wind (g45)", "tstm wind 40",
             "tstm wind 45", "tstm wind and lightning", "tstm wind g45", "tstm wind/hail", "tsunami", "typhoon",
             "unseasonable cold", "unseasonably cold", "unseasonably warm", "unseasonal rain",
             "urban/sml stream fld", "volcanic ash", "warm weather", "waterspout", "wet microburst", "whirlwind",
             "wild/forest fire", "wildfire", "wind", "wind and wave", "wind damage", "winds", "winter storm",
             "winter weather", "winter weather mix", "winter weather/mix", "wintry mix"),

    EVENT=c("frost/freeze", "storm surge/tide", "avalanche", "coastal flood", "ice storm", "blizzard", "dust devil",
             "ice storm", "wildfire", "coastal flood", "coastal flood", "coastal flood", "coastal flood",
             "coastal flood", "marine thunderstorm wind", "marine thunderstorm wind", "cold/wind chill",
             "cold/wind chill", "cold/wind chill", "cold/wind chill", "cold/wind chill", "flash flood",
             "frost/freeze", "dense fog", "thunderstorm wind", "drought", "heavy rain", "drought", "dust devil",
             "dust storm", "frost/freeze", "coastal flood", "excessive heat", "heavy snow",
             "extreme cold/wind chill", "extreme cold/wind chill", "extreme cold/wind chill",
             "extreme cold/wind chill", "heavy snow", "flash flood", "flash flood", "flood", "flash flood",
             "dense fog", "frost/freeze", "frost/freeze", "freezing fog", "winter strom", "freezing fog",
             "frost/freeze", "frost/freeze", "funnel cloud", "frost/freeze", "strong wind", "high wind",
             "thunderstorm wind", "high wind", "thunderstorm wind", "high wind", "hail", "ice storm", "high surf",
             "heat", "heat", "heavy rain", "heavy rain", "high surf", "heavy snow", "heavy snow", "high surf",
             "high surf", "high surf", "high surf", "high surf", "high surf", "high surf", "high surf", "high wind",
             "high wind", "high wind", "hurricane (typhoon)", "hurricane (typhoon)", "hurricane (typhoon)",
             "excessive heat", "heavy snow", "flood", "ice storm", "ice storm", "ice storm", "ice storm",
             "lake-effect snow", "lake-effect snow", "lakeshore flood", "debris flow", "debris flow", "debris flow",
             "debris flow", "winter weather", "freezing fog", "heavy snow", "heavy snow", "lightning",
             "marine high wind", "marine high wind", "marine strong wind", "marine thunderstorm wind",
             "marine thunderstorm wind", "blizzard", "sleet", "heavy rain", "debris flow", "debris flow",
             "debris flow", "high wind", "high wind", "high wind", "heavy rain", "sleet", "excessive heat", "flood",
             "flood", "flood", "flood", "debris flow", "high surf", "high surf", "high surf", "seiche", "hail",
             "heavy snow", "ice storm", "heavy snow", "heavy snow", "storm surge/tide", "storm surge/tide",
             "strong wind", "strong wind", "thunderstorm wind", "thunderstorm wind", "thunderstorm wind",
             "coastal flood", "tornado", "heavy rain", "tropical depression", "tropical storm", "thunderstorm wind",
             "thunderstorm wind", "thunderstorm wind", "thunderstorm wind", "thunderstorm wind",
             "thunderstorm wind", "thunderstorm wind", "thunderstorm wind", "thunderstorm wind",
             "thunderstorm wind", "thunderstorm wind", "tsunami", "typhoon", "extrem cold/wind chill",
             "extrem cold/wind chill", "excessive heat", "heavy rain", "flood", "volcanic ash", "excessive heat",
             "waterspout", "waterspout", "funnel cloud", "wildfire", "wildfire", "high wind", "high wind",
             "strong wind", "high wind", "winter storm", "winter weather", "winter weather", "winter weather",
             "winter weather")

 )
```

We can create the new column `EVENT` by mergin *StormData.raw* and *EVTYPEmap*
including all rows in *StormData.raw*:

```{r merge, cache=TRUE}
StormData.raw <- merge(StormData.raw, EVTYPEmap, all.x=TRUE)
```

Now `EVENT` column contains only valid values:

```{r length-EVENT, cache=TRUE}
length(unique(StormData.raw$EVENT))
```

The only value not assigned is *astronomical low tide*: no *EVTYPE* value has
been found that could match this value.

<br>

### Preparing data for health analysis

Only a subset of the current `StormData.raw` is needed to address health impact:

```{r Storm.health, cache=TRUE}
Storm.health <- subset(StormData.raw, FATALITIES>0 | INJURIES>0,
                       c("EVENT", "FATALITIES", "INJURIES"))
```

<br>

### Preparing data for economic analysis

The same occurs with economic impact:

```{r Storm.economy, cache=TRUE}
Storm.economy <- subset(StormData.raw, PROPDMG>0 | CROPDMG>0,
                        c("EVENT", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP"))
```

In the case of economic analysis we need to convert columns `PROPDMG` and `CROPDMG`
to valid values depending on the 10th power in the corresponding columns
`PROPDMGEXP` and `CROPDMGEXP`. We first *normalize strings* in these columns as
a cleanup procedure:

```{r clean-XROPDMGEXP, cache=TRUE}
Storm.economy$PROPDMGEXP <- str_trim(tolower(Storm.economy$PROPDMGEXP))
Storm.economy$CROPDMGEXP <- str_trim(tolower(Storm.economy$CROPDMGEXP))
```

The cleanup is finished. Magnitude indicators are correctly set:

```{r show-XROPDMGEXP, cache=TRUE}
unique(Storm.economy$PROPDMGEXP)
unique(Storm.economy$CROPDMGEXP)
```

The conversion of columns `PROPDMG` and `CROPDMG` to integer values is as
follows:

```{r convert-XROPDMG, cache=TRUE}
power10 <- data.frame(factor=c("", "k", "m", "b"),
                      magnitude=c(1, 1000, 1000000, 1000000000))

Storm.economy <- merge(x=Storm.economy, y=power10,
                       by.x=as.factor("PROPDMGEXP"), by.y="factor",
                       all.x=TRUE, sort=FALSE)
colnames(Storm.economy)[6] <- "magnitude.prop"
Storm.economy$PROPDMG <- Storm.economy$PROPDMG * Storm.economy$magnitude.prop

Storm.economy <- merge(x=Storm.economy, y=power10,
                       by.x=as.factor("CROPDMGEXP"), by.y="factor",
                       all.x=TRUE, sort=FALSE)
colnames(Storm.economy)[7] <- "magnitude.crop"
Storm.economy$CROPDMG <- Storm.economy$CROPDMG * Storm.economy$magnitude.crop
```

<br>

At this point we have a **tidy data set** to start the analysis of health and
economic impacts of storm events.

<br>

## Data Analysis

Analysis made on the tidy data set is quite obvious: data aggregation on a given
aspect will show the total impact of each event.

### Health Impact

We consider of crucial importance to separate *fatalities* from *injuries*:

```{r fatalities-and-injuries, cache=TRUE}
fatalities <- aggregate(Storm.health$FATALITIES, by=list(Storm.health$EVENT), FUN=sum)
names(fatalities) <- c("EVENT", "total")
head(fatalities)

injuries <- aggregate(Storm.health$INJURIES, by=list(Storm.health$EVENT), FUN=sum)
names(injuries) <- c("EVENT", "total")
head(injuries)
```

### Economic Impact

Conversely, economic impact can be aggregated without separation between property
and crop damage, that should be made on a more focused report. The aggregation
made below makes also a *normalization of totals* to express them as milions of
dollars:

```{r economic-impact, cache=TRUE}
Storm.economy$total <- (Storm.economy$PROPDMG + Storm.economy$CROPDMG)/1000000
economy <- aggregate(Storm.economy$total, by=list(Storm.economy$EVENT), FUN=sum)
names(economy) <- c("EVENT", "total")
head(economy)
```

<br>

## Results

Once the data analysis has been concluded it's time to present final results. The 
result is presented as the top ten events of each impact.

<br>

### Health impact

#### Fatalities

Top ten events that are most harmful with respect to population, resulting in
fatalities:

```{r result-fatalities, cache=TRUE}
fatalities.top10 <- arrange(fatalities, desc(total))[1:10, ]
fatalities.top10

ggplot(fatalities.top10, aes(x=reorder(EVENT, total), y=total)) +
    geom_bar(colour="white", fill="red2", stat="identity") +
    xlab("EVENT") +
    ylab("POPULATION IMPACT - Fatalities") +
    coord_flip()
```

<br>

#### Injuries

Top ten events that are most harmful with respect to population, resulting in
injuries:

```{r result-injuries, cache=TRUE}
injuries.top10 <- arrange(injuries, desc(total))[1:10, ]
injuries.top10

ggplot(injuries.top10, aes(x=reorder(EVENT, total), y=total)) +
    geom_bar(colour="white", fill="orange2", stat="identity") +
    xlab("EVENT") +
    ylab("POPULATION IMPACT - Injuries") +
    coord_flip()
```

<br>

### Economic impact

Top ten events that are most harmful with respect to economic impact, both
crops and properties:

```{r result-economy, cache=TRUE}
economy.top10 <- arrange(economy, desc(total))[1:10, ]
economy.top10

ggplot(economy.top10, aes(x=reorder(EVENT, total), y=total)) +
    geom_bar(colour="white", fill="green4", stat="identity") +
    xlab("EVENT") +
    ylab("ECONOMIC IMPACT  (Millions of dollars)") +
    coord_flip()
```
